---
title: "Reproducible Research: Peer Assessment 1"
author: "Adam S"
output: 
  html_document:
    keep_md: true
---

## Loading and preprocessing the data
```{r}

#load libraries
library(ggplot2)
library(dplyr)
library(doBy)

#read the file
activity = read.csv("activity.csv", na.strings="NA", sep=",", stringsAsFactors=FALSE)

#summarizing data by day
activity_day <- group_by(activity, date)
activity_day_sum <- summarize(activity_day, day_sum = sum(steps, na.rm = TRUE)) 
```

##Histogram of the total number of steps taken each day

```{r, echo=FALSE}
hist(activity_day_sum$day_sum, col = "red", main = "Date", xlab = "Steps", breaks= 50)
```

## What are mean and median total number of steps taken per day?

```{r}
summaryBy(steps ~ date, data = activity, FUN = list(mean, median))
```

#plot of the 5-minute interval (x-axis) and the average number of steps taken, averaged across all days (y-axis)
```{r, echo=FALSE}
plot(activity_interval_mean$interval, activity_interval_mean$mean, type= "l", main = "Average Number of Steps Taken in Intervals", ylab = "Steps", xlab = "Time (Minutes)")
```


##The 5-minute interval that, on average, contains the maximum number of steps
```{r}
activity_interval_mean[which.max(activity_interval_mean$mean),1]
```


#The total number of missing values in the dataset (i.e. the total number of rows with NAs)
```{r}
count(activity[!complete.cases(activity),])
```

##Code to describe and show a strategy for imputing missing data
```{r}
#adding daily mean steps as variable to main table
activity_with_mean_interval <- merge(activity, activity_interval_mean, by.x="interval", by.y="interval", all=TRUE, na.rm=TRUE)

#replacing NAs with mean steps for the day
activity_with_mean_interval$steps[is.na(activity_with_mean_interval$steps)] <- 987654321
activity_with_mean_interval$steps <- ifelse(activity_with_mean_interval$steps == 987654321, activity_with_mean_interval$mean, activity_with_mean_interval$steps) 
```


##Histogram of the total number of steps taken each day after missing values are imputed
```{r}
#summarizing the data by date
activity_mean_interval_day <- group_by(activity_with_mean_interval, date)
activity_mean_interval_day_sum <- summarize(activity_mean_interval_day, day_sum = sum(steps, na.rm = TRUE))
```
```{r, echo=FALSE}
par(mfrow = c(2,1))
hist(activity_mean_interval_day_sum$day_sum, col = "red", main = "Date", xlab = "Steps", breaks= 50)
hist(activity_day_sum$day_sum, col = "red", main = "Date", xlab = "Steps", breaks= 50)
```

##Panel plot comparing the average number of steps taken per 5-minute interval across weekdays and weekends


```{r}
#converting date character field to a date field
activity_with_mean_interval$date = strptime(activity_with_mean_interval$date, "%Y-%m-%d")

#adding new 'day' variable
activity_with_mean_interval$day <- weekdays(activity_with_mean_interval$date, abbreviate = TRUE)

#creating a new variable to categorize days into weekday and weekend days 
attach(activity_with_mean_interval)
activity_with_mean_interval$category[day == "Mon"] <- "Weekday"
activity_with_mean_interval$category[day == "Tue"] <- "Weekday"
activity_with_mean_interval$category[day == "Wed"] <- "Weekday"
activity_with_mean_interval$category[day == "Thu"] <- "Weekday"
activity_with_mean_interval$category[day == "Fri"] <- "Weekday"
activity_with_mean_interval$category[day == "Sat"] <- "Weekend"
activity_with_mean_interval$category[day == "Sun"] <- "Weekend"
detach(activity_with_mean_interval)

#subsetting the data between weekend and weekdays
activity_weekend <- subset(activity_with_mean_interval, category == "Weekend", select=c(interval, steps, mean))
activity_weekday <- subset(activity_with_mean_interval, category == "Weekday", select=c(interval, steps, mean))

#calculate mean and median values per day
#group steps by interval and calculate the mean of steps by interval
activity_weekend_interval <- group_by(activity_weekend, interval)
activity_weekend_interval_mean <- summarize(activity_weekend_interval, mean = mean(steps, na.rm = TRUE))

activity_weekday_interval <- group_by(activity_weekday, interval)
activity_weekday_interval_mean <- summarize(activity_weekday_interval, mean = mean(steps, na.rm = TRUE))
```
```{r, echo=FALSE}
plot(activity_weekend_interval_mean$interval, activity_weekend_interval_mean$mean, type = "l", main="Weekend", sub="", xlab="", ylab="Number of Steps", ylim=c(0, 250))
plot(activity_weekday_interval_mean$interval, activity_weekday_interval_mean$mean, type = "l", main="Weekday", sub="", xlab="Interval", ylab="Number of Steps", ylim=c(0, 250))
```
